---
title: "Analysis of Korkola et al. RPPA data"
author: "Evert Bosdriesz"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    # theme: simplex
    # highlight: kate
    toc: true
---

```{r knitr_init, echo=FALSE, cache=FALSE}
rm(list = ls())
library(knitr)
library(rmdformats)


## Global options
options(max.print = "80")
opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(width = 80)
```

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(here)

source(here("R/graphics-settings.R"))
```

# Summary/Background

## Aim

Here we analyze RPPA data from Korkola et al., PLOS ONE, 2015.
The main aim is to see how strong the signal of targeted inhibition is in cells that are not starved and restimulated.
This should inform the decision for the MCF10A WT/PI3Kmut perturbation experiments we plan to do.

## Experimental setup

 * Panel of 22 cell lines were treated with AKTi, Lapatinib, or the combination.
 * Samples were harvested 0.5, 1, 2, 4, 8, 24, 48 and 72 hours after drug treatment for **RPPA** analysis.
 * Cells were **not** straved prior to drug treatment.
 
# Data analysis

## Data import

Get and clean the data.

```{r load d}
df <- readxl::read_xlsx(here("data/Korkola-etal-rppa/journal.pone.0133219.s001.XLSX"), skip = 7)

df <- df %>% 
  separate(`Sample Name`, into = c("cell_line", "treatment", "time"), sep = "\\)|_") %>% 
  mutate(
    cell_line = str_replace_all(cell_line, "\\s|\\([0-9]*", ""),
    time = as.numeric(str_trim(str_replace(str_replace(time, "hr", ""), "1/2", "0.5"))),
    treatment = factor(treatment, levels = c("DMSO", "Lapatinib", "AKTi", "Lap+AKTi"))
    )

head(df)
```

We restrict analysis to epitopes that are part of the Prot-at-once assay we will use.

```{r df_pao}
df_pao <- df %>% 
  select(cell_line, treatment, time, matches("AKT_pS473|MEK1_2_pS217|GSK3a_b_pS21|p70S6K_pT389|S6_pS235")) %>% 
  gather(epitope, expression, -cell_line, -treatment, -time)
```

First, let's look at the response after 1 hour.

```{r plot 1hr}
ggplot(filter(df_pao, time == "2"), aes(x = cell_line, y = expression, fill = treatment, color = treatment)) + 
  geom_point() +
  labs(x = "Time (hrs)", y = "Normalized, log2-transformed centered expression",
       title = "1 hour after drug treatmen") +
  facet_wrap(~epitope) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Of note: 

  * AKT pS473 shows clear effect of feedback activation upon AKT inhibition in most/all cell lines.

For completeness, also the full time course data.

```{r plot timecourse, fig.height=20, fig.width=10}
ggplot(df_pao, aes(x = time, y = expression, color = treatment)) + 
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(x = "Time (hrs)", y = "Normalized, log2-transformed centered expression",
       title = "RPPA time course of selected epitopes") +
  facet_grid(cell_line~epitope) +
  my_theme
```

Of note:

  * For selection of the timepoint: AKT shows little dynamics after 1 hour.
  * pS6 continues to decrease after 1 hour, but in most cell lines where it decreases, clear onset present after 1 hour.
  * Most other epitopes remain flat. This is not very surprising since e.g. the MAPK pathway is not perturbed. 
  However, when using a restimulation experiment, we might see more dynamics.


# Session Info

```{r sessionInfo}
sessionInfo()
```
