---
title: "Luminex data exploration"
author: "Evert Bosdriesz"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    theme: flatly
    highlight: kate
editor_options:
  markdown:
    wrap: sentence
---



```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print = "80")
opts_chunk$set(
  echo = FALSE, cache = FALSE, prompt = FALSE, tidy = FALSE, comment = NA,
  message = FALSE, warning = FALSE
  )
opts_knit$set(width = 80)
```

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(here)
```

```{r read data, message=FALSE}
wt <- read_tsv(here("results", "perturbations", "wt-luminex.tsv"))
pi3k <- read_tsv(here("results", "perturbations", "pi3k-luminex.tsv"))

source(here("R", "graphics-settings.R"))
```



# Correct for protein amount

Should we correct for the total protein amount?
There were small differences between protein amounts, which we've measured prior to sending the samples.
We used these measurements to correct for different input amounts.

What is the distribution of these?

```{r prot_conc}
prot_conc <- readxl::read_xlsx(
  here("data", "Luminex", "2017-12-08", "171201-perturbation-samples.xlsx"),
  sheet = "BCA PTPN11", range = "A22:K118") %>% 
  select("Treatment" = ...2, "Protein_concentration" = ...6, Sample_number) %>% 
  mutate(
    Cell_line = str_replace(if_else(Sample_number <= 48, "WT", "Mut"), "Mut", "PI3K"),
    Sample_number = as.factor(Sample_number)
    ) %>% 
  filter(Treatment != "Lysis buffer") %>% 
  group_by(Cell_line) %>% 
  mutate(mean_conc = mean(Protein_concentration)) %>% 
  ungroup() %>% 
  mutate(correction_factor = Protein_concentration/mean_conc) 

ggplot(prot_conc, aes(x = correction_factor, fill = Cell_line)) + 
  geom_histogram(bins = 20) +
  theme_minimal() +
  scale_fill_manual(
    name = "",
    values = color_lst,
    labels = label_lst,
    breaks = c("WT", "PI3K")
  ) +
  ggtitle("Distribution of protein-concentration correction factors")
```

There is one "outlier" to left. 
Which is it?

```{r table correction factors}
head(arrange(prot_conc, correction_factor))
```


# Plot ERK and AKT

```{r barcharts}
akt_wt <- ggplot(
  mutate(wt, Sample_number = as.factor(Sample_number)), 
  aes(Sample_number, BioAkt)) + 
  geom_bar(stat = "Identity") +
  labs(title = expression("MCF10A Parental"), y = expression("p-AKT"^S473)) +
  ggthemes::theme_hc()
ggsave(here("figures", "Luminex", "pAKT-wt-bar.pdf"), 
       akt_wt, width = 10, height = 2.5)

erk_wt <- ggplot(
  mutate(wt, Sample_number = as.factor(Sample_number)), 
  aes(Sample_number, ERK1)) + 
  geom_bar(stat = "Identity") +
  labs(title = expression("MCF10A Parental"), y = expression("p-ERK"^{T202/Y204})) +
  ggthemes::theme_hc()
ggsave(here( "figures", "Luminex", "pERK-wt-bar.pdf"), 
       erk_wt, width = 10, height = 2.5)

akt_mut <- ggplot(
  mutate(pi3k, Sample_number = as.factor(Sample_number)), 
  aes(Sample_number, BioAkt)) + 
  geom_bar(stat = "Identity") + 
  labs(title = expression("MCF10A PI3K"^H1074R), y = expression("p-AKT"^S473)) +
  ggthemes::theme_hc()
ggsave(here( "figures", "Luminex", "pAKT-mut-bar.pdf"), 
       akt_mut, width = 10, height = 2.5)

erk_mut <- ggplot(
  mutate(pi3k, Sample_number = as.factor(Sample_number)),
  aes(Sample_number, ERK1)) + 
  geom_bar(stat = "Identity") + 
  labs(title = expression("MCF10A PI3K"^H1074R), y = expression("p-ERK"^{T202/Y204})) +
  theme_minimal()
ggsave(here( "figures", "Luminex", "pERK-mut-bar.pdf"), 
       erk_mut, width = 10, height = 2.5)

akt_wt
akt_mut
erk_wt
erk_mut
rm(akt_wt, akt_mut, erk_wt, erk_mut)
```

# Baseline differences between cell lines

As expected, pAKT and PRAS40 are higher in the PI3K mutant.
RS6 is also slighly elevated.
Replicates look fine for most epitopes, but for some reason the variance is somewhat larger in the WT cells.

```{r plot dmso, message=FALSE}

labels = c("EGFR" = "EGFR Y1068",
          "MEK1" = "MEK1 S217/S221",
          "ERK1" = "ERK1 T202/Y204",
          "GSK3" = "GSK3 S21/S9",
          "CREB1" = "CREB1 S133",
          "BioAkt" = "AKT1 T473",
          "PRAS40" = "PRAS40 T246",
          "P70S6K" = "P70S6K T3890",
          "RS6" = "RPS6 S235")


dmso <- bind_rows(list("WT" = wt,"PI3K" = pi3k), .id = "Cell") %>% 
  filter(Treatment == "DMSO") %>% 
  select(-Sample_number, -Treatment) %>% 
  gather(Epitope, Expression, -Cell) %>% 
  filter(Epitope != "Akt1") %>% 
  mutate(Epitope = factor(Epitope, levels = names(labels)))

p <- ggplot(dmso, aes(x = factor(Cell, levels = c("WT", "PI3K")), y = Expression, color = Cell)) +
  geom_boxplot() +
  expand_limits(y = 0) + 
  labs(title = "Epitope intensity in DMSO (4 replicates)", x = "", y = "Intensity (AU)") +
  scale_color_manual(
    name = "",
    values = color_lst,
    labels = label_lst,
    breaks = c("WT", "PI3K")
  ) +
  scale_x_discrete(labels = list(WT = bquote("Parental"), PI3K = bquote("PI3K"^{H1047R}))) +
  cowplot::theme_half_open(font_family = "Helvetica") +
  theme(legend.position = "none") +
  ggpubr::stat_compare_means(
    method = 't.test', label = 'p.signif', hide.ns = TRUE,
    label.x = 1.5, label.y.npc = 0.8, size = 3, family = "Helvetica", face = "bold")


p + facet_wrap(~Epitope, scales = "free_y", labeller = as_labeller(labels))
  
p_exp <- p + 
  cowplot::theme_half_open(font_family = "Helvetica", font_size = 6) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_wrap(~Epitope, scales = "free_y", labeller = as_labeller(labels)) +
  labs(title = "Baseline signaling activity", x = "", y = "Intensity (AU)") 

ggsave(
  here("figures", "luminex", "DMSO-controls.pdf"), p_exp,
  height = 2.5, width = 2.5
)
#   
# rm(dmso)

```

## Correlation in Log2-fold changes

```{r}
wt_lfc <- read_tsv(here("results", "perturbations", "wt-luminex-lfc.tsv"))
pi3k_lfc <- read_tsv(here("results", "perturbations", "pi3k-luminex-lfc.tsv"))


p <- inner_join(
  pivot_longer(wt_lfc, -Epitope, names_to = "treatent", values_to = "lfc_wt"),
  pivot_longer(pi3k_lfc, -Epitope, names_to = "treatent", values_to = "lfc_pi3k")
  ) %>% 
  filter(Epitope != "Akt1") %>% 
  ggplot(aes(x = lfc_wt, y = lfc_pi3k)) +
  geom_point(size = 0.8) +
  #coord_equal() +
  facet_wrap(~Epitope,  scales = "free", labeller = as_labeller(labels)) +
  labs(title = "Log2 fold change compared to DMSO",
       x = bquote("Parental"), y  = bquote("PI3K"^{H1047R})) +
  cowplot::theme_cowplot(font_family = "Helvetica")

p

p_exp <- p + 
  facet_wrap(~Epitope,  scales = "free", labeller = as_labeller(labels), ncol = 2) +
  cowplot::theme_half_open(font_family = "Helvetica", font_size = 8)

ggsave(
  here("figures", "luminex", "correlation-WT-PI3K.pdf"), p_exp,
  height = 4, width = 3
)

```


# Session Info

```{r sessionInfo}
sessionInfo()
```
