---
title: "The mapping from signaling and drug response"
description: "Parametrizing the relation between short term signaling response and long term cell viability"
output: 
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    theme: flatly
    highlight: kate
editor_options:
  markdown:
    wrap: sentence
---

# Introduction

In our experiments, viabilities are measured in a 3-day CellTiter-Blue assay, while signaling responses are measured in the Luminex 2 hours after drug perturbation. Here, we test to what extend differences in viability can be predicted from the measured signaling response, and parametrize this relation.

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
```

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(here)
library(broom)
library(rmarkdown)

source(here("R/graphics-settings.R"))
```

# Data preparation

The signaling model fits/predicts log2-fold changes in nodes in response to perturbations, so we'll only consider those.

```{r read data, message=FALSE}
wt_lum_lfc <- read_tsv(here("results", "perturbations", "wt-luminex-lfc.tsv"))
pi3k_lum_lfc <- read_tsv(here("results", "perturbations", "pi3k-luminex-lfc.tsv"))

wt_lum_abs <- read_tsv(here("results", "perturbations", "wt-luminex.tsv"))
pi3k_lum_abs <- read_tsv(here("results", "perturbations", "pi3k-luminex.tsv"))

wt_env <- read_tsv(here("results", "perturbations", "wt-envision.tsv"))
pi3k_env <- read_tsv(here("results", "perturbations", "pi3k-envision.tsv"))
```

First, we prepare the data-frame dat\_lfc, containing the log2 fold-changes of each epitope together with the Viabilities. The FGFRi treatments are discarded in the network reconstruction, so we'll also not include it here.

```{r dat_lfc}
wt <- left_join(
  wt_env %>% 
    group_by(Treatment) %>% 
    summarise(Viability = mean(NormalizedViability)),
  wt_lum_lfc %>%
    gather(Treatment, val, -Epitope) %>% 
    spread(Epitope, val),
  by = "Treatment"
)

pi3k <- left_join(
  pi3k_env %>%
    group_by(Treatment) %>%
    summarise(Viability = mean(NormalizedViability)),
  pi3k_lum_lfc %>%
    gather(Treatment, val, -Epitope) %>%
    spread(Epitope, val),
  by = "Treatment"
)

dat_lfc <- bind_rows(list("WT" = wt, "PI3K" = pi3k), .id = "Cell") %>%
  filter(!is.na(Akt1)) %>%
  filter(!str_detect(Treatment, 'FGFRi')) %>% # Remove treatments not used in signaling model
  mutate(Cell = factor(Cell, levels = c("WT", "PI3K")))
rm(wt, pi3k)


paged_table(head(dat_lfc))
```


# Correlations between log2 fold change and viability

A first simple approach is to look at the correlations between Epitopes and Inhibition.

We see a pretty nice negative correlation between the log2-fold change in pAKT and pERK, and drug efficacy. However, none of the nodes is very predictive for viabilty by itself.

```{r figure epitopes-viability-measured-scatter, fig.height=6, message=FALSE}
ggplot(
  filter(
    gather(dat_lfc, Epitope, LFC, -Cell, -Treatment, -Viability),
    Epitope != "Akt1"
  ),
  aes(x = LFC, y = Viability, color = Cell)) +
  geom_vline(xintercept = 0, color = "gray", linetype="dotted") +
  geom_hline(yintercept = 1, color = "gray", linetype="dotted") +
  expand_limits(y = c(0, 1)) +
  geom_point() +
  scale_color_manual(
    name = "",
    breaks = c("WT", "PI3K"),
    values = color_lst,
    labels = label_lst
  ) +
  labs(
    title = "Correlation between signaling responses and viability",
    x = bquote("Log"[2]~"fold-change in epitope"), y = "Viability (Relative to DMSO)"
  ) +
  facet_wrap(~Epitope, scales = "free", labeller = as_labeller(epitope_labels)) +
  cowplot::theme_cowplot(font_family = "Helvetica") +
  theme(legend.position = "bottom")
```

```{r figure delpAKT-viability, message=FALSE, eval=FALSE}
plt <- ggplot(
  filter(
    gather(dat_lfc, Epitope, LFC, -Cell, -Treatment, -Viability),
    Epitope %in% c("BioAkt", "ERK1")
  ),
  aes(x = LFC, y = Viability, color = Cell)) +
  geom_vline(xintercept = 0, color = "gray", linetype="dotted") +
  geom_hline(yintercept = 1, color = "gray", linetype="dotted") +
  expand_limits(y = c(0, 1)) +
  geom_point() +
  labs(
    #title = "Correlation between signaling responses and viability",
    x = bquote("Log"[2]~"fold-change in epitope"), y = "Viability (Relative to DMSO)"
  ) +
  facet_wrap(~Epitope, scales = "free", labeller = as_labeller(epitope_labels)) +
  scale_color_manual(name = "", values = color_lst, labels = label_lst) +
  cowplot::theme_cowplot(font_family = "Helvetica", font_size = 6) +
  theme(legend.position = "bottom")

plt + labs(title = "Correlation between signaling responses and viability")

```

ERK1 and AKT have the strongest correlation with Inhibition.

```{r correlation lfc}
dat_lfc %>%
  summarise_if(is.double, funs(cor(., dat_lfc[, "Viability"]))) %>%
  gather(Epitope, Correlation) %>%
  filter(Epitope != "Viability") %>%
  arrange(desc(Correlation)) %>%
  paged_table()
```
# Michaelis-Menten-like transfer functions

We saw that ∆pAKT1 and ∆pERK1 both had a reasonable correlation with viability, clearly not perfect.
We therefor try a model that incorportates both:
 
$$\mathrm{viability} \sim \frac{1}{1 - \Delta ERK1/KM_{ERK1} - \Delta AKT/KM_{AKT}}$$

```{r fit nonlinear on full data, message=FALSE}
subtitle_mm <- expression(
  "Viability ~ " * frac("1","[1 + -"*Delta*"AKT/KM"[AKT] * " + -"
  * Delta*"ERK1/KM"[ERK] * "]"))

formula_mm  <- Viability ~ (1/(1 + -ERK1 / KM_ERK1 + -BioAkt / KM_BioAkt))

fit_mm_wt  <- nls(
  formula_mm, filter(dat_lfc, Cell == "WT"), 
  start = list("KM_ERK1" = 0.5, "KM_BioAkt" = 2.5)
  )

fit_mm_pi3k  <- nls(
  formula_mm, filter(dat_lfc, Cell == "PI3K"), 
  start = list("KM_ERK1" = 0.5, "KM_BioAkt" = 2.5)
  )

prediction_wt <- broom::augment(fit_mm_wt, filter(dat_lfc, Cell == "WT")) %>% 
  rename(PredictedViability = .fitted)

prediction_pi3k <- broom::augment(fit_mm_pi3k, filter(dat_lfc, Cell == "PI3K")) %>% 
  rename(PredictedViability = .fitted)


prediction_combined <- bind_rows(prediction_wt, prediction_pi3k)


ggplot(prediction_combined,
  aes(x = Viability, y = PredictedViability)) +
  geom_abline(slope = 1, color = "lightgray") +
  geom_point(aes(color = Cell)) +
  labs(
    title = "Correlation between measured and\nfitted viabilty",
    subtitle = subtitle_mm,
    x = "Measured viability", y = "Viability fitted from pERK and pAKT"
  ) +
  #expand_limits(x = 0, y = 0) + 
  ggpubr::stat_cor(method = "pearson") +
  scale_color_manual(name = "", values = color_lst, labels = label_lst) +
  cowplot::theme_cowplot(font_family = "Helvetica") +
  theme(legend.position = "bottom")
# 
# ggsave(here('figures', 'correlations-and-predictions', 'viability-signaling-MM-all.pdf'),
#        width=2, height=2)
```

We obtain the following parameters from this.

```{r write parameters to file}
params <- bind_rows(
  list("WT" = broom::tidy(fit_mm_wt),
       "PI3K" = broom::tidy(fit_mm_pi3k)),
  .id = "Cell"
    
)

params
write_tsv(params,
          here('results', 'perturbations', 'signaling-viability-parameters.tsv')
)
```

# Model assesment using LOO cross-validation

While we have a model that has just to parameters, we nonetheless still check the performance in a leave-one-out cross validation loop.


```{r loo_results, warning=FALSE}
# Function to perform the model fitting
get_loo_results <- function(splits, formula, starts = NA, grouped = FALSE,
                            method = "nls", maxiter = 100){
  # Get the analysis data
  dat <- rsample::analysis(splits)

  # If a grouped (i.e. per-cell-line) model is used, remove analisys data not
  # matching the assesment sample.
  if (grouped) {
    dat <- filter(dat, Cell == rsample::assessment(splits)$Cell)
  }
  # Train the model
  if (method == "nls") {
    mod <- nls(formula = formula, data = dat, start = starts,
               nls.control(maxiter = maxiter))
  } else if (method == "lm") {
    mod <- lm(formula = formula, data = dat)
  }

  # Obtain model parameters and predictions
  bind_cols(
    select(broom::augment(mod, newdata = rsample::assessment(splits)),
           Cell, Treatment, Viability, .fitted),
    spread(select(broom::tidy(mod), term, estimate), term, estimate)
  )
  #  broom::augment(mod, newdata = rsample::assessment(splits))
}

#-------------------------------------------------------------------------------
# MM-like models

loo_mm_grouped <- dat_lfc %>%
  filter(Treatment != "DMSO") %>%
  rsample::loo_cv() %>%
  pull(splits) %>%
  map(get_loo_results,
      formula = formula_mm,
      start = list("KM_ERK1" = 0.8, "KM_BioAkt" = 4),
      grouped = TRUE,
      maxiter = 1000
  ) %>%
  bind_rows() %>%
  mutate(grouping = "per-cell-line", model = "MM") %>%
  rename(PredictedViability = .fitted)

summarise(loo_mm_grouped, pearson = cor(Viability, PredictedViability))
```
As expected, the performance in a cross-validation loop is comparable (but slighlty worse).

# Parameter uncertainty estimation using bootstrapping

The model fits the data fairly well.
To perform parameter uncertainty estimation, and asses to what extend the parameters might differ between the cell lines, we perform 1000 bootstraps.

```{r bootstap models}
boot_wt <- 
  dat_lfc %>%
  filter(Cell == "WT") %>%
  rsample::bootstraps(times = 1000) %>%
  pull(splits) %>%
  map(
    possibly(
      function(split) {
        nls(formula = formula_mm,
            data = rsample::analysis(split),
            start = list("KM_ERK1" = 0.6, "KM_BioAkt" = 3.9)) %>%
          broom::tidy() %>%
          select(term, estimate) %>%
          spread(term, estimate)
      },
      otherwise = tibble(KM_BioAkt = NA, KM_ERK1 = NA)
    )
  ) %>%
  bind_rows(.id = "bootstrap")

boot_pi3k <- 
  dat_lfc %>%
  filter(Cell == "PI3K") %>%
  rsample::bootstraps(times = 1000) %>%
  pull(splits) %>%
  map(
    possibly(
      function(split) {
        nls(formula = formula_mm,
            data = rsample::analysis(split),
            start = list("KM_ERK1" = 0.6, "KM_BioAkt" = 3.9)) %>%
          broom::tidy() %>%
          select(term, estimate) %>%
          spread(term, estimate)
      },
      otherwise = tibble(KM_BioAkt = NA, KM_ERK1 = NA)
    )
  ) %>%
  bind_rows(.id = "bootstrap")

boot_results <- bind_rows(
  list( "WT" = boot_wt, "PI3K" = boot_pi3k),
  .id = "model")
```

The plot below shows the 95% confidence intervals of the parameters overlap between the cell lines. 

```{r plot bootstrapped parameter estimates}
tmp <- gather(boot_results, term, estimate, KM_BioAkt, KM_ERK1) %>%
  group_by(model, term) %>%
  summarise(
    median = median(estimate, na.rm = TRUE),
    mean = mean(estimate, na.rm = TRUE),
    ci_lower = quantile(estimate, probs = 0.025, na.rm = TRUE),
    ci_upper = quantile(estimate, probs = 0.975, na.rm = TRUE)
  )

plt <- ggplot(filter(tmp, model != "combined"), aes(x = model, y = mean, color = model)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = .1) +
  geom_point(size = 3, shape = 21, fill = "white") +
  facet_wrap(~term, scales = "free", ncol = 4) +
  coord_flip() +
  scale_color_manual(name = "", breaks = c("WT", "PI3K"),
                     values = color_lst,
                     labels = label_lst
  ) +
  cowplot::theme_half_open(font_family = "Helvetica") +
  theme(legend.position = "bottom", axis.text.y =element_blank()) +
  labs(y = "Parameter estimate", x = "",
       title = "Parameter estimates and 95% confidence intervals",
       subtitle = "Obtained using 1000 bootstraps"
  )
plt
```

# Conclusion

-   The phenotypic response (Viability) can be predicted well from the measured signaling responses. This is surprising as signaling is measured after 2 hours, whereas viability is a 3-day assay.
-   Changes in ERK1 and AKT1 are most informative.

# Session Info

```{r sessionInfo}
sessionInfo()
```
