---
title: "Analysis of the validation experiments"
output: 
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    theme: flatly
    highlight: kate
editor_options:
  markdown:
    wrap: sentence
---

Here, I analyze the data of validation experiments that we performed to test 3-drug, low-dose combinations that were predicted to be selective for the MCF10A wildtype cells as compared to the PI3K mutants. This experiment was performed twice, first on 15-11-2018 and than again on 03-12-2018 because the first time the cell-count was on the low side. Here, we only analyse the data from 03-12-208, although the other set gives similar results.

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
```

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(here)
library(broom)
library(rmarkdown)
library(patchwork)

source(here("R/graphics-settings.R"))
```

```{r load data, message=FALSE}
df <- read_tsv(here("results", "validation", "20181203-envision.tsv"))
```

The experiment was performed in 8 replicate plates and contained:

-   Low dose 3-drug combinations predicted to be selective (N=`r length(unique(filter(df, type == "Selective")$combo))` combos).
-   Low dose 3-drug combinations predicted to be non-selective (N=`r length(unique(filter(df, type == "NonSelective")$combo))` combos).
-   Single drug controls (low dose, at their IC10)
-   DMSO and PAO as negative and positive control, respectively (N=8 and N=10 per cell line per plate).

The data have been normalized using the script [20181203-wrangle-envision.R](https://github.com/evertbosdriesz/mcf10a-pi3k/src/master/R/validation/20181203-wrangle-envision.R).

We normalized on a per-plate, per-cell line basis as the parental cell line has an overall somewhat slower proliferation rate. Overall, the replicates look reasonably tight.

```{r plot_normalization}
p1 <- 
  ggplot(
  filter(df, combo %in% c("DMSO", "PAO")),
  aes(x = replicate, y = viability)
  ) +
  geom_boxplot(aes(color = cell_line)) +
  scale_color_manual(
    name = "",
    values = color_lst,
    labels = label_lst
  ) +
  labs(
    title = "Non-normalized controls",
    x = "Replicate plate",
    y = "Viability [AU]"
  ) +
  facet_wrap(~combo) +
  theme_light(base_family = "Helvetica") +
  theme(legend.position = "bottom") 

p2 <- ggplot(
  filter(df, combo %in% c("DMSO", "PAO")),
  aes(x = replicate, y = normalized_viability)
  ) +
  geom_boxplot(aes(color = cell_line)) +
  scale_color_manual(
    name = "",
    values = color_lst,
    labels = label_lst
  ) +
  labs(
    title = "Normalized controls",
    x = "Replicate plate",
    y = "Viability [AU]"
  ) +
  facet_wrap(~combo) +
  theme_light(base_family = "Helvetica") +
  theme(legend.position = "bottom")

p1 + p2
```

The most important test of these validation experiments is if our method can prioritize selective combinations. The box-plot below shows that **combinations that are predicted to be selective are indeed more selective than the control, non-selective combinations** and thus validates model predictions.

```{r selective vs nonselective all, message=FALSE}
df_sel <- df %>%
  filter(type %in% c("Selective", "NonSelective")) %>%
  select(replicate, cell_line, combo, type, normalized_viability, treatment) %>%
  spread(cell_line, normalized_viability) %>%
  mutate(selectivity = PI3K - WT) %>% 
  group_by(combo, treatment, type) %>%
  summarise(selectivity = mean(selectivity))

plt_validation <- 
  ggplot(df_sel, aes(x = type, y = selectivity, color = type)) +
  geom_boxplot(outlier.colour = NA) +
  labs(
    x = "",
    y = "Measured selectivity"
  ) +
  scale_color_manual(values = c("gray", "red")) +
  scale_x_discrete(labels = c("Control", "Predicted\nSelective")) +
  cowplot::theme_cowplot(font_family = "Helvetica", font_size = 6) +
  theme(legend.position = "none")
  
plt_validation  + 
  geom_jitter(width=0.25, size=0.5) +
  #ggbeeswarm::geom_beeswarm() +
  cowplot::theme_cowplot(font_family = "Helvetica", font_size = 14) +
  theme(aspect.ratio = 1, legend.position = "none") +
  ggpubr::stat_compare_means(method = "wilcox.test", label.x = 1.25)  
  

```

We can also perform a t-test for each individual combination to asses if it is selective.

```{r}
ttest_combo <- 
  df %>%
  filter(type == "Selective") %>%
  select(treatment, cell_line, normalized_viability) %>%
  group_by(treatment) %>% 
  mutate(is_partental = cell_line == "WT") %>% 
  do(tidy(t.test(.$normalized_viability ~ .$cell_line, alternative = "greater"))) %>% 
  arrange(p.value)

ttest_combo %>% 
  paged_table()

write_tsv(ttest_combo, here("results", "validation", "ttest-combos.tsv"))
```
How many of these are significant? ```r filter(ttest_combo, p.value < 0.05) %>% nrow()```

```{r message=FALSE}
predictions_control <-
  read_tsv(here("results", "perturbations", "optimization",
                "optimization-selection-control.tsv")) %>%
  select(treatment = AppliedConcentrations,
         WT_predicted = WT,
         PI3K_predicted = PI3K,
         selectivity_predicted = Selectivity) %>%
  mutate(type = "NonSelective")

predictions_selective <-
  read_tsv(here("results", "perturbations", "optimization",
                "optimization-selection-wtselective.tsv")) %>%
  select(treatment = AppliedConcentrations,
         WT_predicted = WT,
         PI3K_predicted = PI3K,
         selectivity_predicted = Selectivity) %>%
  mutate(type = "Selective")

predictions <- bind_rows(predictions_control, predictions_selective)

```

Since IGF1Ri is also selective as a single treatment (cf. pilot data), this raises the question what the added value of the combining it with other drugs is. In other words, are the IGF1Ri + 2-other drugs combinations more selective than IFG1Ri alone?

Let's first test this over all IGF1Ri containing combinations (and all replicates).

```{r df_igf1r}
df_igf1r <- df %>%
 filter(
  (n_drugs == 3 & str_detect(combo, "IGF1Ri") & type == "Selective") |
    (n_drugs == 1 & str_detect(combo, "IGF1Ri") & IGF1Ri > 1) # Only use the higher control concentration
  ) %>%
  select(type, replicate, cell_line, combo, type, normalized_viability, treatment) %>%
  spread(cell_line, normalized_viability) %>%
  mutate(selectivity = PI3K - WT)

group_by(df_igf1r, treatment, type) %>% 
  summarise(selectivity = mean(selectivity))
```

```{r boxplot_df_igf1r}
igf1r_mono_sel <- mean(filter(df_igf1r, treatment == "IGF1Ri")$selectivity)

plt_igf1ri_treatments <- 
  ggplot(
    filter(df_igf1r, type == "Selective") %>% 
      group_by(type, treatment) %>% 
      summarise(selectivity = mean(selectivity)),
  aes(x = type, y = selectivity)
  ) +
  geom_boxplot(outlier.colour = NA, color = "red", width=0.5) +
  geom_hline(yintercept = igf1r_mono_sel, color = "gray") +
  labs(
    #title = "All IGF1Ri-containing treatments",
    x = "",
    y = "Measured selectivity"
  ) +
  scale_x_discrete(
    labels = c("SingleDrugControl" = "IGF1Ri\nonly",  "Selective" = "IGF1Ri +\n2 other drugs"),
    ) 

plt_igf1ri_treatments +
  geom_jitter(width = 0.25, size = 0.5, color = "red") +
  cowplot::theme_cowplot(font_family = "Helvetica", font_size = 14) +
  theme(legend.position = "None") +
  annotate(geom="text", x = 1, y=igf1r_mono_sel + 0.0125, label="IGF1R\nmonotreatment", color="gray")

```
What is the one-samnple t-test comparing IGF1Ri alone to IGF1R in combo?

```{r}
tmp <- df_igf1r %>% 
  group_by(type, treatment) %>% 
  summarise(WT = mean(WT), PI3K = mean(PI3K)) %>% 
  mutate(selectivity = PI3K - WT)

t.test(
  filter(tmp, type == "Selective")$selectivity,
  mu = filter(tmp, type == "SingleDrugControl")$selectivity,
  alternative = "greater"
  )
```

The boxplot above clearly indicates that **IGF1Ri in combination with selected other drugs is more selective than IGF1Ri alone.**

However, in this case it makes more sense to look at a per-combinations basis. (Now, we use a one-sided t-test instead of wilcoxon because we're not combining different combinations, only comparing replicates.)

```{r ttest_combos}
ttest_igf1r_combos <- df_igf1r %>% 
  filter(type != "SingleDrugControl") %>% 
  left_join(
    select(
      filter(df_igf1r, type == "SingleDrugControl"),
      replicate, control_selectivity = selectivity
      ),
    by = "replicate"
  ) %>% 
  group_by(combo) %>%
  do(tidy(t.test(.$selectivity, .$control_selectivity, alternative  = "greater"))) %>% 
  ungroup() %>% 
  arrange(p.value) %>% 
  mutate(rank = dense_rank(desc(estimate)))

ttest_igf1r_combos %>% 
  select(combo, estimate, p.value) %>% 
  paged_table()

write_tsv(ttest_igf1r_combos, here("results", "validation", "ttest-igf1ri-mono-vs-combo.tsv"))
```

**`r nrow(filter(ttest_igf1r_combos, p.value < 0.05))` out of `r nrow(ttest_igf1r_combos)` IGF1Ri-containing combinations individually show a statistically significant greater selectivity than IGF1Ri treatment alone** at a comparable concentration.

For completeness, we also check that other low-dose monotherapies do not induce any selectivity.

```{r plot all single drug comtrols}
df %>% 
  filter(type == "SingleDrugControl") %>% 
  ggplot(aes(x = combo, y = normalized_viability, color = cell_line)) +
  geom_boxplot() +
  scale_color_manual(
    name = "", breaks = c("WT", "PI3K"),
    values = color_lst,
    labels = label_lst
  ) +
  labs(
    title = "Single drug controls",
    y = "Viability (Relative to DMSO)",
    x = ""
  ) +
  cowplot::theme_cowplot(font_family = "Helvetica", font_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_compare_means(
    method = "wilcox.test", label.x = 1.5, label.y = 1.1, 
    size = 3, label = 'p.signif') 
```

Finally, let's visualize the most selective combinations.

```{r plot_top_combos, message=FALSE}
combos_to_plot <- 
  ttest_igf1r_combos %>% 
  arrange(desc(conf.low)) %>% 
  #head(n = 8) %>% 
  pull(combo)

background_col <- "gray90"

plt_top_combos <- 
  df_igf1r %>% 
  filter(combo %in% c("IGF1Ri", combos_to_plot)) %>% 
  left_join(ttest_igf1r_combos, by = "combo") %>% 
  select(replicate, combo, PI3K, WT, rank) %>% 
  gather(cell_line, viability, WT, PI3K) %>% 
  ggplot(
  aes(x = combo, y = viability, color = cell_line)
  ) +
  geom_rect(xmin=0.5, xmax=1.5, ymin=-Inf, ymax=+Inf, fill = "gray", color = "gray") +
  geom_rect(xmin=2.5, xmax=3.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=4.5, xmax=5.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=6.5, xmax=7.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=8.5, xmax=9.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=10.5, xmax=11.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=12.5, xmax=13.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=14.5, xmax=15.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_rect(xmin=16.5, xmax=17.5, ymin=-Inf, ymax=+Inf, fill = background_col, color = background_col) +
  geom_boxplot(outlier.size = 0.5) +
  scale_x_discrete(limits = c("IGF1Ri", combos_to_plot)) +
  scale_color_manual(
    name = "", breaks = c("WT", "PI3K"),
    values = color_lst,
    labels = label_lst
  ) +
  labs(
    #title = "Selective combinations",
    y = "Viability (Relative to DMSO)",
    x = ""
  ) +
  cowplot::theme_cowplot(font_family = "Helvetica") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

plt_top_combos
```

# Session info

```{r session_info}
sessioninfo::session_info()
```


